---
title: "Tooth Growth Analysis"
author: "Ramana Sonti"
date: "2/5/2017"
output: pdf_document
---

### Synopsis:
###### This report analyzes experimental data reported on tooth growth with increasing doses of vitamin C and orange joice administered to a set of 60 guinea pigs. It uses T confidence intervals and hypothesis testing to conclude that both VC and OJ help improve tooth growth with increasing doses.

#### Data Description:
###### Each of 60 guinea pigs were given one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid, a form of vitamin C and coded as VC).

```{r setup, include = FALSE, echo = FALSE}
Sys.setenv(cache.path = "cache")
knitr::opts_chunk$set(cache=TRUE)

# helper function to install/load the packages if they are missing
usePackage <- function(p) {
    if (!is.element(p, installed.packages()[,1]))
        install.packages(p, dep = TRUE)
    require(p, character.only = TRUE)
}
```
### Exploratory Data Analysis:
##### Load the ToothGrowth data and perform some basic exploratory data analyses
```{r echo = FALSE}
usePackage("datasets")
usePackage("pastecs")
usePackage("stats")
usePackage("R.utils")
usePackage("dtplyr")
usePackage("ggplot2")
usePackage("grid")
usePackage("reshape2")

data("ToothGrowth")
head(ToothGrowth)
str(ToothGrowth)
g <- ggplot(ToothGrowth, aes(x = dose, y = len, color = supp, size = 1)) + geom_point() + geom_line()
g
#g <- ggplot(ToothGrowth, aes(x = dose, y = len, dose = factor(supp)))
#g <- g + geom_line(size = 1, aes(colour = supp)) + geom_point(size =10, pch = 21, fill = "salmon", alpha = .5)
#g
```
### Confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose:
```{r echo = FALSE}
vctg05<- subset(ToothGrowth, supp=="VC" & dose == 0.5)[,1]
vctg10<- subset(ToothGrowth, supp=="VC" & dose == 1.0)[,1] 
vctg20<- subset(ToothGrowth, supp=="VC" & dose == 2.0)[,1] 
ojtg05<- subset(ToothGrowth, supp=="OJ" & dose == 0.5)[,1] 
ojtg10<- subset(ToothGrowth, supp=="OJ" & dose == 1.0)[,1]
ojtg20<- subset(ToothGrowth, supp=="OJ" & dose == 2.0)[,1]
```
### Hypothesis Testing for VC:
##### Null Hypothesis HO: VC has no effect on tooth growth
##### Alternative Hypothesis Ha: Tooth growth increases with increasing levels of VC.
```{r echo = TRUE}
t.test(vctg10, vctg05, paired = FALSE, var.equal = FALSE)
t.test(vctg20, vctg05, paired = FALSE, var.equal = FALSE)
```
### Hypothesis Testing for OJ: 
##### Null Hypothesis HO: OJ has no effect on tooth growth 
##### Alternative Hypothesis Ha: Tooth growth increases with increasing levels of OJ.
```{r echo = TRUE}
t.test(ojtg10, ojtg05, paired = FALSE, var.equal = FALSE)
t.test(ojtg20, ojtg05, paired = FALSE, var.equal = FALSE)
```
### Conclusion:
For both VC and Oj, null hypothesis has been rejected as the t-test results show that the sample mean corrsponding to higher dose falls outside of 95% confidence intrval.