---
title: "Weather Events with Major Impact on Public Health and Economy"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
Sys.setenv(cache.path = "cache")
knitr::opts_chunk$set(cache=TRUE)

# helper function to install/load the packages if they are missing
usePackage <- function(p) {
    if (!is.element(p, installed.packages()[,1]))
        install.packages(p, dep = TRUE)
    require(p, character.only = TRUE)
}

```


### Synopsis: 
##### The weather data from NOAA for the years between 1950 and 2011 has been analysed to identify the the events that caused the most damage to the economy and the public health across the United States of America. The initial dataset has been subjected to a set of procedures to address some inconsistencies in the way the event types were recorded. A set of tests have been performed to confirm that the new dataset produced after the cleanup does not require any further cleanup to determine the events that caused the most damage. Out of 476 billion dollars of total economic damage from all events, the damage caused by the floods with about 180 billion dollars tobs the list followed by the damage casued by hurricanes and tornadoes. The tornadoes were responsible for the most number of fatalities and injuries.

## Data Processing:
##### 1. Enable cache on all chunks to avoid redoing the steps if the objects did not change 
##### 2. Download StormData.csv.bz2 fomr url if it doesn't exist in the current directory
##### 3. Uncompress the file using bunzip2 if it is not uncompressed already
##### 4. Read StormData.csv into stormdata data frame
##### 5. Check summary of stormdata to understand the soruce data before transformation
##### 6. Build dbgexpkv refrence data frame with damage exponent and its value
##### 7. Create stormdata.econ by subsetting stormdata with observations that do not have zeros for both PROPDMG and CROPDMG
#####    and selecing EVTYPE and PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP columns
##### 8. Merge dbgexpkv into stormdata.econ to add exponent value column 
##### 9. Create stormdata.econdmg by mutating EVTYPE to replce '/' with ' ' and to convert all event types to upper case.
##### 10. Calculate total damage for each type, aggregate on EVTYPE, and arrange ECONDMG in descending order
##### 11. Display summary on stormdata.econdmg and create a barplot for top 30 event types
#####  
##### 12. Identify key words representing the events from top of the barplot above.
#####     FLOOD/FLOODING HURRICANE TORNADOES WIND/WND THUNDERSTORM/TSTM WINT/WINTER
##### 13. Aggregate on EVTYPE with above keywords to get individual total damage
##### 14. Compare above totals against the top 6 on the barplot
##### 
##### 15. Create stormdata.health by subsetting stormdata with observations that do not have zeros for both FATALITIES and INJURIES
#####     and selecting EVTYPE, FATALITIES, and INJURIES columsn
##### 16. Aggregate FATALITIES and INJURIES by EVTYPE and arrange FATALITIES in descending order
##### 17. Display summary and create a bar plot for the top 20 events
##### 18. Calculate number of FATALITIES and INJURIES for each event type identified in step 12.
##### 19. Build stormdata.top6 data frame from ECONDMG from step 13 and FATALITIES/INJURIES from step 18
##### 20. Create stormdata.top6.econ and stormdata.top6.health from stormdata.top6
##### 21. Create bar plots from stormdata.top6.health and stormdata.top6.econ on the same grid.

#### Loading and Preprocessing the Data:
``` {r echo=TRUE, cache = TRUE}
usePackage("stats")
usePackage("R.utils")
usePackage("dtplyr")
usePackage("ggplot2")
usePackage("grid")

# source
sourceURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
sourceFileCompressed<-"StormData.csv.bz2"
sourceFile<-"StormData.csv"

start_time <- Sys.time()

# downlod the compressed file if it doesn't exist in the current directory
if (!file.exists(sourceFileCompressed)) {
  download.file(sourceURL, destfile = sourceFileCompressed, mode = "wb")
  print(paste0("It took ", Sys.time() - start_time, " to download ", sourceFileCompressed, "."), quote = FALSE)
}

start_time <- Sys.time()

# if not uncompressed already
if (!file.exists(sourceFile)) {
  bunzip2(sourceFileCompressed, overwrite = T, remove = FALSE)
  print(paste0("It took ", Sys.time() - start_time, " to uncompress ", sourceFileCompressed, "."), quote = FALSE)
}

start_time <- Sys.time()

# read the file into a data frame
if (file.exists(sourceFile)) {
    stormdata <- read.csv(sourceFile, sep = ",", header = TRUE)
    # print "Skip reading the file..."
    print(paste0("It took ", Sys.time() - start_time, " to read ", sourceFile, "."), quote = FALSE)
} else {
  print(paste0(sourceFile, " does not exist!"), quote = FALSE)
  exit (-1)
}

```


## Data Transformation and Analysis:
``` {r echo = TRUE, cache = TRUE}
ls()
summary(stormdata)
library(grid)

#  Based on  the articale "How To Handle Exponent Value of PROPDMGEXP and CROPDMGEXP" at
# https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html
# create reference table for exponents used and the corresponding multipliers
dmgexpkv = data.table(x=c("", "-", "?", "+", 0, seq(1:8), "H", "h", "K","k","M","m","B","b")
                      , y=c(0,0,0,1,rep(10,9),100,100,1000,1000,1000000,1000000,1000000000,1000000000))
names(dmgexpkv) <- c("PROPDMGEXP", "PROPDMGEXPVAL")
#

# Consider observarions for which both PROPDMG and CROPDMG are not 0
stormdata.econ <- subset(stormdata, !(PROPDMG==0 & CROPDMG==0), select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
#
# merge to get PROPDBMGEXPVAL column added to stormdata.econ
stormdata.econ <- merge(stormdata.econ, dmgexpkv, by.x = "PROPDMGEXP", by.y = "PROPDMGEXP")
#
# merge to get CROPDBMGEXPVAL column added to stormdata.econ
names(dmgexpkv) <- c("CROPDMGEXP", "CROPDMGEXPVAL")
stormdata.econ <- merge(stormdata.econ, dmgexpkv, by.x = "CROPDMGEXP", by.y = "CROPDMGEXP")
#
# Calculate total damage for each event
stormdata.econdmg <- stormdata.econ %>% mutate(EVTYPE = gsub("/", " ", toupper(EVTYPE)), ECONDMG = PROPDMG*PROPDMGEXPVAL+CROPDMG*CROPDMGEXPVAL) %>% select(EVTYPE,ECONDMG)
# sum/aggreate the damage by EVTYPE
stormdata.econdmg <- aggregate(ECONDMG ~ EVTYPE, data = stormdata.econdmg, FUN = 'sum') %>% arrange(desc(ECONDMG))
#
dim(stormdata.econdmg)
summary(stormdata.econdmg)

#
# have enough left margin for event type lables
par(mar = c(5,12,4,1) + 0.1)
# plot economic damage in billions of dollars for top 30 events 
barplot(as.matrix(head(stormdata.econdmg$ECONDMG/1000000000,30))
        , beside = TRUE
        , col=rainbow(5)
        , horiz = TRUE
        , names.arg = head(stormdata.econdmg$EVTYPE,30)
        , cex.names = 0.8
        , las = 1
        , xlab = "Economic Damage (billions of dollars)" 
        , main = "Top 30 events with greatest economic consequences"
        , xpd = TRUE)

# Check totals by event types that have the keywords associated with most common major severe events
# FLOOD/FLOODING
# HURRICANE
# TORNADOES
# WIND/WND
# THUNDERSTORM/TSTM
# WINT/WINTER
#
flood <- round(sum(stormdata.econdmg[grep("FLOOD", stormdata.econdmg$EVTYPE),2])/1000000000)
hurric <- round(sum(stormdata.econdmg[grep("HURRIC", stormdata.econdmg$EVTYPE),2])/1000000000)
tornado <- round(sum(stormdata.econdmg[grep("TORN", stormdata.econdmg$EVTYPE),2])/1000000000)
wind <- round(sum(stormdata.econdmg[grep("WIND|WND", stormdata.econdmg$EVTYPE),2])/1000000000)
tstm <- round(sum(stormdata.econdmg[grep("THUNDERSTORM|TSTM|THUNDER", stormdata.econdmg$EVTYPE),2])/1000000000)
winter <- round(sum(stormdata.econdmg[grep("WINTER", stormdata.econdmg$EVTYPE),2])/1000000000)

print("Total damage in billions of dollars for the top 6 event catagories :", quote = FALSE)
cat(  "floods           :", flood
    , "\nhurricanes       : ", hurric
    , "\ntornadoes        : ", tornado
    , "\nwinds            : ", wind
    , "\nthunderstorms    : ", tstm
    , "\nwinter           :  ", winter)

### consider observarions for which both PROPDMGEXP and CROPDMGEXP are not 0
# 
stormdata.health <- subset(stormdata, !(FATALITIES==0 & INJURIES==0), select = c(EVTYPE, FATALITIES, INJURIES)) %>% mutate (EVTYPE = gsub("/", " ", toupper(EVTYPE)))

stormdata.health <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data = stormdata.health, FUN = 'sum') %>% arrange(desc(FATALITIES), desc(INJURIES))

# summary
summary(stormdata.health)

stormdata.health.matrix <- t(head(stormdata.health,20)[-1])
colnames(stormdata.health.matrix) <- head(stormdata.health,20)[, 1]
# plot health damage for top 30 events 
barplot(stormdata.health.matrix
        , beside = TRUE
        , col=rainbow(2)
        , horiz = TRUE
        , names.arg = colnames(stormdata.health.matrix)
        , cex.names = 0.8
        , las = 1
        , xlab = "Number of Fatalities/Injuries" 
        , main = "Top 20 most harmful events to population health"
        , xpd = TRUE)

legend("topright", c("FATALITIES", "INJURIES"), cex=1.0, bty="n", fill=rainbow(2))


# 
flood.f <- sum(stormdata.health[grep("FLOOD", stormdata.health$EVTYPE),2])
flood.i <- sum(stormdata.health[grep("FLOOD", stormdata.health$EVTYPE),3])
hurric.f <- sum(stormdata.health[grep("HURRIC", stormdata.health$EVTYPE),2])
hurric.i <- sum(stormdata.health[grep("HURRIC", stormdata.health$EVTYPE),3])
tornado.f <- sum(stormdata.health[grep("TORN", stormdata.health$EVTYPE),2])
tornado.i <- sum(stormdata.health[grep("TORN", stormdata.health$EVTYPE),3])
wind.f <- sum(stormdata.health[grep("WIND|WND", stormdata.health$EVTYPE),2])
wind.i <- sum(stormdata.health[grep("WIND|WND", stormdata.health$EVTYPE),3])
tstm.f <- sum(stormdata.health[grep("THUNDERSTORM|TSTM|THUNDER", stormdata.health$EVTYPE),2])
tstm.i <- sum(stormdata.health[grep("THUNDERSTORM|TSTM|THUNDER", stormdata.health$EVTYPE),3])
winter.f <- sum(stormdata.health[grep("WINTER", stormdata.health$EVTYPE),2])
winter.i <- sum(stormdata.health[grep("WINTER", stormdata.health$EVTYPE),3])

stormdata.top6 <- data.frame(c("FLOODS", "HURRICANES", "TORNADOES", "WINDS", "THUNDERSTORMS", "WINTER")
                   ,c(flood, hurric, tornado, wind, tstm, winter)
                   ,c(flood.f, hurric.f, tornado.f, wind.f, tstm.f, winter.f)
                   ,c(flood.i, hurric.i, tornado.i, wind.i, tstm.i, winter.i))

names(stormdata.top6) <- c("EVTYPE", "ECONDMG", "FATALITIES", "INJURIES")

stormdata.top6.health <- subset(stormdata.top6, select = c(EVTYPE, FATALITIES, INJURIES))
stormdata.top6.econ <- subset(stormdata.top6, select = c(EVTYPE, ECONDMG))

stormdata.top6.health.melt <- melt(stormdata.top6.health)
stormdata.top6.econ.melt <- melt(stormdata.top6.econ)
names(stormdata.top6.health.melt) <- c("EVTYPE", "DAMAGE", "DAMAGEVAL")
names(stormdata.top6.econ.melt) <- c("EVTYPE", "DAMAGE", "DAMAGEVAL")

plot1 <- 
  ggplot(stormdata.top6.health.melt) +
  geom_bar(aes(x = EVTYPE , y = DAMAGEVAL, fill = DAMAGE), stat="identity", position="dodge") +
  labs(title = "Damage caused by top 6 events") + 
  ylab("Number of Fatalities/Injured") +
  theme(axis.title.y=element_text(margin=margin(0,20,0,0))) +
  theme(axis.text.x=element_text(size=6, color="red")) +
  theme(panel.margin.y = unit(10, "lines")) +
  theme(axis.title.x = element_blank())

plot2 <- 
  ggplot(stormdata.top6.econ.melt) +
  geom_bar(aes(x = EVTYPE, y = DAMAGEVAL, fill = DAMAGE), stat="identity") +
  ylab("Damage to Economy \n (billions of dollors)") +
  theme(panel.margin.y = unit(10, "lines")) +
  theme(axis.title.y=element_text(margin=margin(0,20,0,0))) +
  theme(axis.text.x=element_blank()) +
  theme(axis.title.x = element_blank())

grid.newpage()
grid.draw(rbind(ggplotGrob(plot1), ggplotGrob(plot2), size = "last"))

  
```

## Analysis:
#### The first plot depicts the economic damage estimates for the top 30 event types. Further cleaning of the data to fix typos and other issues proved unnecessary as the damage contributed by the rest of the event types will not alter the position/rank of the top 5 event types.
#### The second plot depicts the number of fatalities and injured for the top 20 event types. The counts for the rest of the event types are negligible. 
#### The third plot depicts the economoic damage and the fatalities/injuries casued by the top 6 event types. They are shown one above the other in the same grid for comparison.

## Results:
#### The analysis finds that FLOODING causes the most economic damage (~180 billion dollars) of all event tyeps where as tornodoes cause the most damage to the humans (5661 fatalities/91407 injuries).
